Future cosmic microwave background delensing with galaxy surveys.

### 2017-07-26 Notes/galaxy_delensing.tex
abby comments

\def\l{\ell}
% For various journals
%\newcommand{\apj}{ApJ}
\newcommand{\physrep}{Physics Reports}
\newcommand{\jcap}{JCAP}
\newcommand{\apjl}{ApJL}
\def\araa{\ref@jnl{ARA\&A}}
\newcommand{\apjs}{ApJS}
\newcommand{\aap}{A\&A}
\newcommand{\mnras}{MNRAS}
\newcommand{\physrev}{Phys. Rev.}
\newcommand{\physrevlett}{Phys. Rev. Lett.}
\newcommand*\aj{AJ}
%\newcommand*\prd{Phys.~Rev.~D}

@@

\section{Introduction}
\label{sec:intro}
In the standard cosmological paradigm the early universe underwent a period of near-exponential expansion called ``cosmic inflation''.
All the cosmological observations agree with this picture, making it a compelling and elegant description of the Universe initial conditions. Despite the experimental effort, other possible explanations are still valid, and a conclusive evidence of inflation is still to be found. 
Inflation generically predicts a stochastic background of gravitational waves \cite[see e.g.][for a review]{kamionkowski15}. This prediction sets inflation apart from other theories and a detection of primordial gravitational waves could be the compelling evidence cosmologists are looking for. 
These primordial gravitational waves in the early universe would imprint a unique signature on the polarized anisotropies of the CMB. For this reason, CMB polarization is a promising dataset to understand the physics of the early universe and ultimately test inflation.
In particular, we can decompose the CMB polarization fields in Fourier space into even-parity (divergence) and odd-parity (curl) components, referred to as ``E'' and ``B'' modes.
 In the standard scenario, the B-mode polarization is a clean probe of primordial gravitational waves, because these are the only source of B-modes at the epoch of recombination.

Because the CMB B-modes provide the cleanest known observational window into the primordial gravitational wave background, improving their measurement is a major objective of current and future CMB experiments.
Even if inflationary B-modes have not been detected yet, the target value $r \gtrsim 10^{-3}$, should be reachable shortly given the level of noise expected in future CMB experiments \citep{kamionkowski15,abazajian:2016}.
However, just reducing the level of noise will not be enough to attain this goal.
Indeed, the observed B modes are not solely sourced by early universe physics; they are also produced by secondary effects taking place in the late-time low-redshift universe. 
In this work, we will focus on the effect of gravitational interactions with large scale structures (LSS).
Lensing shears the CMB polarization pattern, producing ``lensing B modes'' from CMB E modes \citep{zaldarriaga98}.
This expected component has now been measured both in cross-correlation with LSS \citep{hanson13,bicep2-collaboration:2016,polarbear2014c,van-engelen:2015,planck2015XV} and from CMB data alone \citep{polarbear2014b,bicep2a,keisler15}.

@@

\begin{table}
\caption{$\alpha$: Improvements on $\sigma(r)$ due to delensing for S3 experiments. Here we define LSS-S3 as the combination of S2 LSS tracers (LSS-S2) and DESI. We also use a CMB S3 experiment for the lensing internal reconstruction. See \refsec{S3} for details.}
%The absolute error value of $\sigma(r)$ after delensing for $f_{sky} = $ is $\sigma(r) =$ 
\hspace{2.2cm}
\begin{tabular}{|c | c | c | c|}
\hline
Surveys & $\alpha_{r=0}, N^{B}_{\ell}=0$ & $\alpha_{r=0}$ & $\alpha_{r=0.12}$ \\ \hline \hline
DESI (10\%) &1.12 & 1.08  & 1.03 \\ \hline
LSS-S3 (40\%) &1.73& 1.42  & 1.2\\ \hline
CMB 3G (56\%) &2& 1.57  & 1.21 \\ \hline
LSS+CMB (69\%)& 2.8 & 1.8& 1.31  \\ \hline
\end{tabular}
\label{tab:S3}
\end{table}

@@

The correlation factor with CMB lensing for Stage-4 experiments is shown in \reffig{corrs4_10}  and in \reffig{corrs4_01} for SKA1 and SKA2 respectively.
%=======
\begin{figure}[htbp]
\begin{center}
\includegraphics[scale=1.]{../images/S4_scenario_ska10.pdf}
\caption{Correlation factor. Same as \reffig{corrnow} but for stage 4 experiments. }
\label{fig:corrs4_10}
\end{center}
\end{figure}
%=======
\begin{figure}[htbp]
\begin{center}
\includegraphics[scale=1.]{../images/S4_scenario_ska01.pdf}
\caption{Correlation factor. Same as \reffig{corrnow} but for stage 4 experiments. }
\label{fig:corrs4_01}
\end{center}
\end{figure}
%=======
A tomographically binned LSST-like experiment will be a very efficient CMB lensing tracer. Indeed it will be more than 70\% correlated with CMB lensing for all the scales $\ell<600$, a performance similar to a Stage-3 CMB internal reconstruction.
A very similar level of efficiency will be achieved by a SKA1-like experiment.
With SKA and LSST, for the first time we will have LSS tracers with a higher delensing efficiency than the CIB.
Despite the improvement in galaxy surveys, the CMB internal reconstruction will still be the main source of delensing. Having a perfect kernel overlap with the true lensing potential it will benefit from the very low level of noise of CMB S4 experiments.
\reffig{corrs4_10} shows that a CMB-S4 experiment will be able to internally reconstruct the lensing potential at more than 90\% up to $\ell=400$.
At this level, CMB S4 internal delensing will not be limited by noise but by small secondary effects like foregrounds contamination, filtered modes etc.
Finally, \reffig{corrs4_01} shows that a very futuristic SKA-2 experiment will have a very high correlation with CMB lensing, being more than 80\% correlated at all the scales. This performance is comparable to a CMB-S4 internal reconstruction.

\reftab{S4} shows the improvement on the inflationary constraint given the cross-correlation factors described above. 
LSST alone will be able to remove almost half of the B-mode lensing power. This will lead to an improvement in $\sigma(r)$ of a factor of 1.8. If we add all the other infrared and optical LSS tracers previously included (i.e. excluding SKA) the improvement factor increase to 2.3.
These gives an important opportunity: for sufficiently high values of $r$ it will be possible to confirm a detection of primordial gravity waves using spectra delensed with LSS, completely independent of CMB internal data. 
Similarly to \cite{namikawa:2016a} we find that SKA phase-1 radio survey will remove 53\% of the lensing power and, once combined with optical and infrared surveys, this will allow the level of power removed using LSS (62\%) to surpass the level of the CMB S3 internal reconstruction.
Finally, as previously mentioned, CMB S4 will be the main source of delensing with the ability of removing up to more than 80\% of the B-modes power. This will bring an almost factor of 4 improvement in the null test case even with realistic noise. Combining with LSS will only increase the removed power compared to CMB only by $\simeq 5\%$ but it will be highly beneficial to test the robustness of the final result.

@@

\begin{table}
\caption{$\alpha$: Gen 4 experiments}
\begin{tabular}{|c | c | c | c|}
\hline
Surveys & $\alpha_{r=0}, N^{B}_{\ell}=0$ & $\alpha_{r=0}$ & $\alpha_{r=0.12}$ \\ \hline \hline
LSST (48\%) &1.88 & 1.81  &  \\ \hline
LSS  (60\%) &2.47 & 2.31  & \\ \hline
SKA (10$\mu$Jy) (53\%) &2.18& 2.06  & \\ \hline
SKA (0.1$\mu$Jy) (77\%) &4.6& 3.96  & \\ \hline
LSS + SKA (10$\mu$Jy)  (62\%) &2.74 & 2.53  &  \\ \hline
LSS + SKA (0.1$\mu$Jy)  (80\%) &4.66 & 3.96  &  \\ \hline
CMB S4 (81\%) & 4.63 & 3.96  &  \\ \hline
LSS+CMB (86\%)& 5.2 &  &  \\ \hline
\end{tabular}
\label{tab:S4}
\end{table}
%=======================================
%\begin{table}
%\caption{$\alpha$: Gen 4 experiments}
%  \begin{tabular}{|c | c | c | c|}
%\hline
%Surveys & $\alpha_{r=0}, N^{B}_{\ell}=0$ & $\alpha_{r=0}$ & $\alpha_{r=0.12}$ \\ \hline \hline
%LSST (10\%) &1.12 & 1.08  & 1.03 \\ \hline
%SKA (0.1$\mu$Jy) (40\%) &1.73& 1.42  & 1.2\\ \hline
%CMB S4 (56\%) &2& 1.57  & 1.21 \\ \hline
%LSS+CMB (69\%)& 2.8 & 1.8& 1.31  \\ \hline
%\end{tabular}
%\label{tab:S4_2}
%\end{table}
%=======================================
%\subsection{Bias uncertainties degradation}
%The uncertainties in the theoretical assumptions used to model the galaxies can cause a degradation of the improvement of the inflationary constraint of delensed spectra.
%In this section, we quantify this effect.
%We will now marginalize over unknown galaxies parameters, but we will use a full dataset of CMB and galaxies data. The idea is that as shown in the low level of noise in galaxy surveys might allow us to calibrate them internally.
%We will use a Fisher approach the Fisher matrix is:
%\beqn
%F_{pq} &=& \sum_{l_{a} =l^{BB}_{\mathrm{min}}}^{l^{BB}_{\mathrm{max}}}  \sum_{l_{b} =l^{BB}_{\mathrm{min}}}^{l^{BB}_{\mathrm{max}}}   \frac{\partial {C}_{l_a}^{BB,\mathrm{del}}}{\partial \theta_p } \left[ \mathrm{Cov}^{BB,BB}\right]_{l_a, l_b} ^{-1} \frac{\partial {C}_{l_b}^{BB,\mathrm{del}}}{\partial \theta_q }\nonumber \\&+&  \sum_j \frac{\frac{\partial {C}_j^{\kappa I}}{\partial \theta_p } \frac{\partial {C}_j^{\kappa I}}{\partial \theta_q }}{(\Delta C_j^{\kappa I})^2}
%+\sum_j \frac{\frac{\partial {C}_j^{I I}}{\partial \theta_p } \frac{\partial {C}_j^{I I}}{\partial \theta_q }}{(\Delta C_j^{I  I})^2}
%\eeqn
%
%\beq
%\alpha_{\mathrm{marginalized}} = \sigma_0(r) / \sigma_{\mathrm{marginalized/delensed}}(r)
%\eeq
%
%
%where the parameter array contains both the tensor to scalar rate $\theta=r$, and the galaxy surveys parameters like the bias $b_i$ or $p_i$ \footnote{In our analysis, since the fiducial value of $r$ is zero,
%the derivative of $C_\ell^{\rm BB,res}$ is non-zero if $\theta_i=r$.}.
%
%We compute the derivatives of the power spectra as described
\section{Conclusions}
\label{sec:concl}
The ability to separate the lensing component of the CMB B-modes from a possible primordial inflationary signal (``delensing'') is necessary to test inflation using the next generation of CMB polarization experiments. 
To delens, we need to accurately reconstruct the large scale structures that lens the CMB in order to build a template of the expected lensing B-modes in the observed patch.
In this paper, we studied the potential impact of large-scale structure galaxy surveys in this important endeavor.
%For current experiment, the cosmic infrared background had already proven to be very efficient in delensing \cite{larsen:2016,manzotti:2017}.

@@

\bibliographystyle{apsrev4-1}
\bibliography{/Users/alessandromanzotti/Work/Astrophysics/latex_bib/cosmobib,/Users/alessandromanzotti/Work/Astrophysics/delensing/sptpol_papers/2016/delens100d/delens100d,/Users/alessandromanzotti/Work/Astrophysics/delensing/sptpol_papers/BIBTEX/spt}
%/Users/alessandromanzotti/Work/Astrophysics/delensing/sptpol_papers/BIBTEX/spt}

>>>>

\def\l{\ell}
% For various journals
%\newcommand{\apj}{ApJ}
\newcommand{\physrep}{Physics Reports}
\newcommand{\jcap}{JCAP}
\newcommand{\apjl}{ApJL}
\newcommand{\araa}{ARA\&A}
\newcommand{\apjs}{ApJS}
\newcommand{\aap}{A\&A}
\newcommand{\mnras}{MNRAS}
\newcommand{\physrev}{Phys. Rev.}
\newcommand{\physrevlett}{Phys. Rev. Lett.}
\newcommand*\aj{AJ}
%\newcommand*\prd{Phys.~Rev.~D}

@@

\section{Introduction}
\label{sec:intro}
In the standard cosmological paradigm the early universe underwent a period of near-exponential expansion called ``cosmic inflation''.
All the cosmological observations agree with this picture, making it a compelling and elegant description of the Universe initial conditions. Despite the experimental effort, other possible explanations are still valid, and a conclusive evidence of inflation is still to be found. 
Inflation generically predicts a stochastic background of gravitational waves \cite[see e.g.][for a review]{kamionkowski:2016}. This prediction sets inflation apart from other theories and a detection of primordial gravitational waves could be the compelling evidence cosmologists are looking for. 
These primordial gravitational waves in the early universe would imprint a unique signature on the polarized anisotropies of the CMB. For this reason, CMB polarization is a promising dataset to understand the physics of the early universe and ultimately test inflation.
In particular, we can decompose the CMB polarization fields in Fourier space into even-parity (divergence) and odd-parity (curl) components, referred to as ``E'' and ``B'' modes.
 In the standard scenario, the B-mode polarization is a clean probe of primordial gravitational waves, because these are the only source of B-modes at the epoch of recombination.

Because the CMB B-modes provide the cleanest known observational window into the primordial gravitational wave background, improving their measurement is a major objective of current and future CMB experiments.
Even if inflationary B-modes have not been detected yet, the target value $r \gtrsim 10^{-3}$, should be reachable shortly given the level of noise expected in future CMB experiments \citep{kamionkowski:2016,abazajian:2016}.
However, just reducing the level of noise will not be enough to attain this goal.
Indeed, the observed B modes are not solely sourced by early universe physics; they are also produced by secondary effects taking place in the late-time low-redshift universe. 
In this work, we will focus on the effect of gravitational interactions with large scale structures (LSS).
Lensing shears the CMB polarization pattern, producing ``lensing B modes'' from CMB E modes \citep{zaldarriaga98}.
This expected component has now been measured both in cross-correlation with LSS \citep{hanson13,bicep2-collaboration:2016,polarbear2014c,van-engelen:2015,planck2015XV} and from CMB data alone \citep{polarbear2014b,bicep2a,keisler15}.

@@

\begin{table}
\caption{$\alpha$: Improvements on $\sigma(r)$ due to delensing for S3 experiments. Here we define LSS-S3 as the combination of S2 LSS tracers (LSS-S2) and DESI. We also use a CMB S3 experiment for the lensing internal reconstruction. See \refsec{S3} for details.}
%The absolute error value of $\sigma(r)$ after delensing for $f_{sky} = $ is $\sigma(r) =$ 
\hspace{2.2cm}
\begin{tabular}{|c | c | c | c|}
\hline
Surveys & $\alpha_{r=0}, N^{B}_{\ell}=0$ & $\alpha_{r=0}$ & $\alpha_{r=0.12}$ \\ \hline \hline
DESI (10\%) &1.12 & 1.08  & 1.03 \\ \hline
LSS-S3 (40\%) &1.73& 1.42  & 1.2\\ \hline
CMB 3G (56\%) &2& 1.57  & 1.21 \\ \hline
LSS-S3+CMB (69\%)& 2.8 & 1.8& 1.31  \\ \hline
\end{tabular}
\label{tab:S3}
\end{table}

@@

The correlation factor with CMB lensing for Stage-4 experiments is shown in \reffig{corrs4_10}  and in \reffig{corrs4_01} for SKA1 and SKA2 respectively.
%=======
\begin{figure}[htbp]
\begin{center}
\includegraphics[scale=1.]{../images/S4_scenario_ska10.pdf}
\caption{Correlation factor $\rho$ with the lensing potential $\kappa$ as a function of the angular scale $\ell$. This figure is the same as \reffig{corrnow} but with a stage-4 CMB experiment performing the internal reconstrcution and with LSST and SKA added to the stage-2 and stage-3 galaxy surveys (LSS-S3). }
\label{fig:corrs4_10}
\end{center}
\end{figure}
%=======
\begin{figure}[htbp]
\begin{center}
\includegraphics[scale=1.]{../images/S4_scenario_ska01.pdf}
\caption{Correlation factor $\rho$ with the lensing potential $\kappa$ as a function of the angular scale $\ell$. This figure is the same as \reffig{corrnow} but with a stage-4 CMB experiment performing the internal reconstrcution and with LSST and SKA added to the stage-2 and stage-3 galaxy surveys (LSS-S3). }
\label{fig:corrs4_01}
\end{center}
\end{figure}
%=======
A tomographically binned LSST-like experiment will be a very efficient CMB lensing tracer. Indeed it will be more than 70\% correlated with CMB lensing for all the scales $\ell<600$, a performance similar to a Stage-3 CMB internal reconstruction.
A very similar level of efficiency will be achieved by a SKA1-like experiment.
With SKA and LSST, for the first time we will have LSS tracers with a higher delensing efficiency than the CIB.
Despite the improvement in galaxy surveys, the CMB internal reconstruction will still be the main source of delensing. Having a perfect kernel overlap with the true lensing potential it will benefit from the very low level of noise of CMB S4 experiments.
\reffig{corrs4_10} shows that a CMB-S4 experiment will be able to internally reconstruct the lensing potential at more than 90\% up to $\ell=400$.
At this level, CMB S4 internal delensing will not be limited by noise but by small secondary effects like foregrounds contamination, filtered modes etc.
Finally, \reffig{corrs4_01} shows that a very futuristic SKA-2 experiment will have a very high correlation with CMB lensing, being more than 80\% correlated at all the scales. This performance is comparable to a CMB-S4 internal reconstruction.

\reftab{S4} shows the improvement on the inflationary constraint given the cross-correlation factors described above. 
LSST alone will be able to remove almost half of the B-mode lensing power. This will lead to an improvement in $\sigma(r)$ of a factor of 1.8. If we add all the other infrared and optical LSS tracers previously included (i.e. excluding SKA) the improvement factor increase to 2.3.
These gives an important opportunity: for sufficiently high values of $r$ it will be possible to confirm a detection of primordial gravitational waves using spectra delensed with LSS, completely independent of CMB internal data. 
Similarly to \cite{namikawa:2016a} we find that SKA phase-1 radio survey will remove 53\% of the lensing power and, once combined with optical and infrared surveys, this will allow the level of power removed using LSS (62\%) to surpass the level of the CMB S3 internal reconstruction.
Finally, as previously mentioned, CMB S4 will be the main source of delensing with the ability of removing up to more than 80\% of the B-modes power. This will bring an almost factor of 4 improvement in the null test case even with realistic noise. Combining with LSS will only increase the removed power compared to CMB only by $\simeq 5\%$ but it will be highly beneficial to test the robustness of the final result.

@@

\begin{table}
\caption{$\alpha$: Gen 4 experiments}
\begin{tabular}{|c | c | c | c|}
\hline
Surveys & $\alpha_{r=0}, N^{B}_{\ell}=0$ & $\alpha_{r=0}$ & $\alpha_{r=0.12}$ \\ \hline \hline
LSST (48\%) &1.84 & 1.76  & 1.74  \\ \hline
LSS-S4  (60\%) &2.47 & 2.31  & \\ \hline
SKA (10$\mu$Jy) (53\%) &2.18& 2.06  & 2.02\\ \hline
SKA (0.1$\mu$Jy) (77\%) &4.6& 3.96  & 3.9\\ \hline
LSS-S4 + SKA (10$\mu$Jy)  (62\%) &2.74 & 2.53  &  \\ \hline
LSS-S4 + SKA (0.1$\mu$Jy)  (80\%) &4.66 & 3.96  &  \\ \hline
CMB S4 (81\%) & 5.16 & 4.32  & 4.2 \\ \hline
LSS+CMB (86\%)& 5.2 &  &  \\ \hline
\end{tabular}
\label{tab:S4}
\end{table}
%=======================================
%\begin{table}
%\caption{$\alpha$: Gen 4 experiments}
%  \begin{tabular}{|c | c | c | c|}
%\hline
%Surveys & $\alpha_{r=0}, N^{B}_{\ell}=0$ & $\alpha_{r=0}$ & $\alpha_{r=0.12}$ \\ \hline \hline
%LSST (10\%) &1.12 & 1.08  & 1.03 \\ \hline
%SKA (0.1$\mu$Jy) (40\%) &1.73& 1.42  & 1.2\\ \hline
%CMB S4 (56\%) &2& 1.57  & 1.21 \\ \hline
%LSS+CMB (69\%)& 2.8 & 1.8& 1.31  \\ \hline
%\end{tabular}
%\label{tab:S4_2}
%\end{table}
%=======================================
%\subsection{Bias uncertainties degradation}
%The uncertainties in the theoretical assumptions used to model the galaxies can cause a degradation of the improvement of the inflationary constraint of delensed spectra.
%In this section, we quantify this effect.
%We will now marginalize over unknown galaxies parameters, but we will use a full dataset of CMB and galaxies data. The idea is that as shown in the low level of noise in galaxy surveys might allow us to calibrate them internally.
%We will use a Fisher approach the Fisher matrix is:
%\beqn
%F_{pq} &=& \sum_{l_{a} =l^{BB}_{\mathrm{min}}}^{l^{BB}_{\mathrm{max}}}  \sum_{l_{b} =l^{BB}_{\mathrm{min}}}^{l^{BB}_{\mathrm{max}}}   \frac{\partial {C}_{l_a}^{BB,\mathrm{del}}}{\partial \theta_p } \left[ \mathrm{Cov}^{BB,BB}\right]_{l_a, l_b} ^{-1} \frac{\partial {C}_{l_b}^{BB,\mathrm{del}}}{\partial \theta_q }\nonumber \\&+&  \sum_j \frac{\frac{\partial {C}_j^{\kappa I}}{\partial \theta_p } \frac{\partial {C}_j^{\kappa I}}{\partial \theta_q }}{(\Delta C_j^{\kappa I})^2}
%+\sum_j \frac{\frac{\partial {C}_j^{I I}}{\partial \theta_p } \frac{\partial {C}_j^{I I}}{\partial \theta_q }}{(\Delta C_j^{I  I})^2}
%\eeqn
%
%\beq
%\alpha_{\mathrm{marginalized}} = \sigma_0(r) / \sigma_{\mathrm{marginalized/delensed}}(r)
%\eeq
%
%
%where the parameter array contains both the tensor to scalar rate $\theta=r$, and the galaxy surveys parameters like the bias $b_i$ or $p_i$ \footnote{In our analysis, since the fiducial value of $r$ is zero,
%the derivative of $C_\ell^{\rm BB,res}$ is non-zero if $\theta_i=r$.}.
%
%We compute the derivatives of the power spectra as described
\section{Conclusions}
\label{sec:concl}
The ability to separate the lensing component of the CMB B-modes from a possible primordial inflationary signal (``delensing'') is necessary to test inflation using the next generation of CMB polarization experiments. 
To delens, we need to accurately reconstruct the large scale structures that lens the CMB in order to build a template of the expected lensing B-modes in the observed patch.
In this paper, we studied the potential impact of large-scale structure galaxy surveys in this important endeavor.
%For current experiment, the cosmic infrared background had already proven to be very efficient in delensing \cite{larsen:2016,manzotti:2017}.

@@

\bibliographystyle{mnrasunsrt}
\bibliography{/Users/alessandromanzotti/Work/Astrophysics/latex_bib/cosmobib,/Users/alessandromanzotti/Work/Astrophysics/delensing/sptpol_papers/2016/delens100d/delens100d,/Users/alessandromanzotti/Work/Astrophysics/delensing/sptpol_papers/BIBTEX/spt}
%/Users/alessandromanzotti/Work/Astrophysics/delensing/sptpol_papers/BIBTEX/spt}
